The effects of alignment error and alignment filtering on the \sw detection of positive selection

### 2011-07-19 projects/slrsim/paper/slrsim.tex
Manuscript revs

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}

\usepackage[a4paper]{geometry}

@@

\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\em Drosophila}\xspace}

@@

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce errors resulting from misalignment, but neither the
prevalence of such errors nor the ability of filters to reduce them
has been systematically studied. Focusing on the sitewise detection of
positive selection, we performed simulation experiments to quantify
the false positives and false negatives introduced by alignment error
and the ability of alignment filters to improve performance. We found
that some aligners produced many false positive results across a wide
range of divergence levels, while others produced very few. False
negatives were a problem for all aligners, increasing with
divergence. Of the aligners tested, PRANK consistently performed the
best and ClustalW performed the worst. The best alignment filters
showed good ability to reduce the error rates from ClustalW
alignments, but they failed to improve the performance of PRANK
alignments under most circumstances. Of the filters tested, GUIDANCE
performed the best and Gblocks performed the worst, failing to improve
even the error-prone ClustalW alignments. The performance of the true
alignment and an optimal filtering method suggested that performance
improvements could be gained by improving aligners or filters to
reduce the prevalence of false negatives, especially at higher
divergence levels and indel rates. We conclude that for the sitewise
evolutionary analysis of most vertebrate, insect and fungal
phylogenies, the use of unfiltered PRANK alignments is recommended.

@@

Furthermore, the evolutionary divergence of a dataset affects the
power of \sw inference and the prevalence of indels in different
ways. As maximum-likelihood methods for detecting positive selection
require data in the form of observed substitution events, they show
little power at low divergence and their highest power at intermediate
to high divergence levels \citep{Anisimova2001Accuracy}. However,
alignment error should be greatest at high divergences, which may have
the effect of reducing power. These two trends suggest that the
overall power will be low at both extremes of divergence, with little
inference power at low divergence (due to the scarcity of data in the
form of observed substitutions) and an overwhelming amount of
alignment error at high divergence (due to the large number of indel
events).

@@

A number of empirical analyses have established that errors in gene
sequencing, annotation and alignment can contribute to errors in
downstream evolutionary analyses such as phylogeny inference
\citep{Wong2008Alignment} et and estimates of positive selection
\citep{Schneider2009Estimates,Markova-Raina2011High} . Most recently,
Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High} showed
that the detection of positively-selected sites and genes in \Dr
genomes is highly sensitive to aligner choice, with PRANK's codon
model consistently producing alignments with the lowest amount of
positive selection. Still, according to the authors' manual inspection
of alignments, even positively-selected sites identified with PRANK
alignments contained a sizable proportion of apparent false positives.

A limitation to the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
used their expected general effect of alignment error (an increase in
false positives due to misalignment of non-homologous codons) as a
proxy by which to compare different methods, allowing for the
conclusion that PRANK was the least error-prone aligner in their
analysis. However, the absolute number of false positives remained
uncertain and there was the possibility of conflating multiple sources
of error: in addition to alignment error, the authors noted that gene
mis-annotation was responsible for many apparent false positives, and
there is also an expected error rate from the likelihood inference
method itself. This limitation leaves important and interesting
questions regarding the nature of alignment error and its quantitative
impact on the detection of positive selection left unanswered by
empirical studies.

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK \citep{Loytynoja2005From}
produce estimates of alignment confidence or reliability. Alignment
confidence scores based on the robustness of alignments to guide tree
perturbations \citep{Penn2010Alignment} or the sampling of suboptimal
alignments \citep{Kim2011PSAR} have also been proposed. Unfortunately,
despite their widespread use, the impact of the many available
filtering methods on phylogenetic and evolutionary analyses has not
been well studied. Even for a single filtering program, Gblocks,
results have been contradictory: one simulation-based study found that
it improved the phylogenetic signal \citep{Talavera2007Improvement}
while an empirical study across a wide range of taxa found that
Gblocks-filtered alignments produced worse phylogenetic trees than
unfiltered alignments \citep{Dessimoz2010Phylogenetic}. With the
application of published filtering methods to alignments before
testing for positive selection becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary estimates. Our approach is similar to that
of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in that we
use simulated protein alignments to evaluate methods for detecting \sw
positive selection, considering each site to be an independent
hypothesis test producing a positive or negative result. Our
experimental design also shares some features with Fletcher and Yang's
\citeyearpar{Fletcher2010Effect} investigation of the performance of
branch-site method for detecting genewise positive selection under
alignment error. Like Fletcher and Yang
\citeyearpar{Fletcher2010Effect}, we simulate codon alignments with
indels and calculate inferred alignments using a small but diverse
sample of aligners. Unlike their study, here we focus on \sw detection
of positive selection occurring throughout a phylogeny (as opposed to
genewise detection of selection acting at specific branches), we
explore a wider range of plausible tree sizes, indel rates and
divergence levels, and we evaluate the impact of a number of alignment
filtering methods on the \sw analysis. Given the recent proliferation
of completed and planned vertebrate genomes, we pay special attention
to choosing simulation parameters and analysis methods similar to
those commonly encountered in the \sw analysis of vertebrate gene
families.

@@

An overview of the simulation parameters used in this study can be
found in Table \ref{table_1}. Three rooted trees were used to guide
the simulation of protein-coding DNA alignments: the artificial
6-taxon tree used by Anisimova and Yang
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate B-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution}, and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure \ref{fig_1}, were scaled to
comparable divergence levels by normalizing their mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.05 and 2.0 synonymous substitutions per synonymous site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table
\ref{table_2}).
\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.5]{fig1.pdf}
\includegraphics[scale=0.8]{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis. The
  original scale for each tree is indicated by a scale bar, but trees
  were scaled to equal mean path length (MPL) divergence levels for
  simulation. (a) A 6-taxon artificial tree used in previous
  simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
  estimated from $\beta$-globin genes of 17 vertebrates and used in
  previous empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
  44-species tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes indicated by
  arrows were used as the reference species when comparing the true
  and inferred alignment (see Methods).}
\label{fig_1}
\end{figure}

The Indelible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons, $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4, and
100 replicate alignments were simulated for each set of conditions
unless otherwise noted. Indel lengths were drawn from a discretized
power-law distribution with an exponential decay parameter of 1.8 and
a maximum value of 40, yielding a mean indel length of 3.33 codons and
standard deviation of 5.51 codons. This power-law model of indel
lengths is well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and manual
inspection of alignments from a range of parameter values identified
the chosen model parameters as resulting in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was scaled between 0 to 0.2 indel events per substitution
per site.

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(unpublished data; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
discrete model of \sw $\omg$ variation was used to approximate the
lognormal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were multiplied by a
correction factor before simulation to correct for the difference
between our definition of branch lengths as the number of synonymous
substitutions per synonymous site ($dS$) and Indelible's
interpretation of branch length as the average number of substitutions
per codon ($t$) \citep{Fletcher2010Effect}. They are related
approximately by $t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and
$S$ are the proportion of nonsynonymous and synonymous sites and
$\bar{\omg}$ is the mean $\omg$ across all sites. $S$ is approximately 0.3
when $\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio for our chosen
distribution is 0.277, yielding a $dS$-to-$t$ conversion factor of
1.48 for all simulations performed.

@@

Alignments were filtered by masking out residues based on the output
of three alignment scoring methods: Gblocks conserved blocks
\citep{Castresana2000Selection}, T-Coffee consistency scores
\citep{Notredame2000TCoffee,Notredame2003Using}, and GUIDANCE
alignment confidence scores \citep{Penn2010Alignment}. Gblocks, which
identifies entire alignment columns as conserved or not conserved, was
run using an increased gap tolerance and a reduced minimum block
length in order to reduce the amount of each alignment removed
(command-line parameters {\em b5=a} and {\em b4=3}), and all residues
from any columns not within an identified conserved block were masked
with $N$s. GUIDANCE and T-Coffee, which produce individual scores for
each residue, were handled differently. A cutoff threshold was chosen
for each method (0.5 for GUIDANCE and 5 for T-Coffee) and residues
equal to or below that threshold were masked. If the cutoff threshold
caused greater than 50\% of an alignment's residues to be masked, the
threshold was relaxed to the highest value for which at least 50\% of
residues remained unmasked. This cap on the severity of each filter
was necessary because the scores from both methods were strongly
affected by the simulation conditions, with much lower average scores
at higher indel rates and divergences. GUIDANCE was run with 100
bootstrap replicates and MAFFT as the bootstrap aligner (except when
\pranka{} or \prankc{} was the original aligner, in which case 30
bootstrap replicates were generated with \pranka{}) and T-Coffee was
run using default settings and the {\em evaluate\_mode
  -output=score\_ascii} command-line parameters to output alignment
scores.

Two unrealistic datasets were produced to serve as controls. First,
the true simulated alignment was included in order to evaluate the \sw
performance without any alignment error. Second, an additional
filtering method was constructed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue reflecting how correctly
is has been placed in the inferred alignment. The approach taken was
to calculate, for each residue, the branch length of the correct
sub-tree (defined as the sub-tree connecting all sequences to which
the current residue was correctly aligned) divided by the branch
length of the total aligned sub-tree (defined as the sub-tree
connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
reflects the expectation that correctly-aligned evolutionary branch
length is the main source of information from which \sw inference
methods derive their power. We refer to this method as the optimal
method. Scores were handled in a manner similar to GUIDANCE and
T-Coffee, using a score threshold of 0.5.

@@

\Sw estimates of selective pressures were calculated using
maximum-likelihood methods implemented in the Phylogenetic Analysis by
Maximum Likelihood (PAML, \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
(SLR, \citealt{Massingham2005Detecting}) software packages.

@@

SLR implements a method specifically designed for \sw estimates which
has been shown in simulations to perform as well as or better than
PAML's \sw random sites models \citep{Massingham2005Detecting}. SLR
models codon evolution as a continuous-time Markov process where
substitutions at one site are independent of substitutions at all
other sites. Importantly, no assumptions are made regarding the
distribution of $\omg$ ratios within the alignment. The value of
$\omg$ is considered to be an independent parameter at each site:
after first optimizing shared parameters using the whole alignment,
SLR uses the shared parameters and the data at each alignment site to
calculate a \sw statistic for non-neutral evolution. This statistic is
based on a likelihood-ratio test where the null model is neutral
evolution ($\omg=1$) and the alternative model is either purifying or
positive selection ($\omg<1$ or $\omg>1$, respectively). The raw
statistic measures the strength of evidence for non-neutral evolution
at each site; following Massingham
\citeyearpar{Massingham2005Detecting} we use a signed version of the
SLR statistic (created by negating the statistic for sites with
$\omg<1$) as the test statistic for positive selection.

@@

In order to compare \sw estimates from different alignments, a single
sequence from each tree was chosen as the reference (arrows, Figure
\ref{fig_1}) and all \sw statistics were mapped from alignment columns
to sequence positions in the reference sequence. This approach bears
some resemblance to the process of mapping alignment-based
evolutionary estimates onto a single member of the alignment for
further analysis and integration with other genome-referenced data (as
is often done, for example, using mammalian alignments and a human
reference). As a result of this reference sequence based mapping,
sites which were deleted in the reference sequence or inserted in a
lineage not ancestral to the reference were dropped from the analysis.

To evaluate the power and error rates that might be achieved in
real-world data analysis, the recommended cutoff thresholds for PAML’s
Bayesian posterior probabilities and the SLR statistic were used to
identify positively selected sites. A posterior probability threshold
of 0.95 was used for PAML \citep{Yang2005Bayes} and a threshold of 3.84,
the 95\% critical value of the $\chi^2$ distribution with 1 degree
of freedom, was used for SLR \citep{Massingham2005Detecting}. Sites were
compared to their true simulated state (e.g. positively-selected or
non-positively selected) in order to identify correct and incorrect
inferences, and from these classifications we calculated the false
positive rate (FPR, defined as the proportion of all sites with true
$\omg<1$ falsely identified as positively selected) and true positive rate
(TPR, defined as the proportion of all sites with true $\omg>1$ correctly
identified as positively selected).

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.75]{fig2.pdf}
\includegraphics[scale=1.25]{fig2.pdf}
\end{center}
\caption{Alignments were simulated without indels for three tree
  shapes and analyzed with SLR, PAML M8, or PAML M2. Fifty replicate
  alignments were simulated for each data point. The performance of
  each analysis method, as measured by four summary statistics, is
  plotted as a function of the mean path length (MPL) divergence. From
  top to bottom: true positive rate (TPR) at the recommended cutoff
  threshold (0.95 for PAML and 3.84 for SLR); false positive rate
  (FPR) at the recommended cutoff threshold; TPR at a 1\% FPR
  threshold; Pearson's correlation coefficient between the true and
  inferred sitewise $\omg$.}
\label{fig_2}
\end{figure}

\section*{Results and Discussion}
\subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
We first evaluated the ability of three \sw methods, PAML M2,
PAML M8 and SLR, to accurately estimate \sw $\omg$ values and to
detect positive selection under a range of tree lengths in the absence
of alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw $\omg$
correlation over a range of MPLs for each of the three simulation
trees.

>>>>

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper]{geometry}
\usepackage{color}

\definecolor{red}{rgb}{1,0,0}

@@

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\em Drosophila}\xspace}

@@

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce errors resulting from misalignment, but neither the
prevalence of such errors nor the ability of filters to reduce them
has been systematically studied. Focusing on the sitewise detection of
positive selection, we performed simulation experiments to quantify
the false positives and false negatives introduced by alignment error
and the ability of alignment filters to improve performance. We found
that some aligners led to many false positive results across a wide
range of divergence levels, while others resulted in very few. False
negatives were a problem for all aligners, increasing with
divergence. Of the aligners tested, PRANK consistently performed the
best and ClustalW performed the worst. The best alignment filters
showed good ability to reduce the error rates from ClustalW
alignments, but they failed to improve the performance of PRANK
alignments under most circumstances. Of the filters tested, GUIDANCE
performed the best and Gblocks performed the worst, failing to improve
even the error-prone ClustalW alignments. The performance of the true
alignment and an optimal filtering method suggested that performance
improvements could be gained by improving aligners or filters to
reduce the prevalence of false negatives, especially at higher
divergence levels and indel rates. \textcolor{red}{As no combination
  of aligner plus filter significantly outperformed the best aligner
  alone,} we conclude that for the sitewise evolutionary analysis of
most vertebrate, insect and fungal phylogenies, the use of unfiltered
PRANK alignments is recommended.

@@

Furthermore, the evolutionary divergence of a dataset affects the
power of \sw inference and the prevalence of indels in multiple
ways. As maximum-likelihood methods for detecting positive selection
require data in the form of observed substitution events, they show
little power at low divergence and their highest power at intermediate
to high divergence levels \citep{Anisimova2001Accuracy}. However,
alignment error should be greatest at high divergences, which may have
the effect of reducing power. These two trends suggest that the
overall power will be low at both extremes of divergence, with little
inference power at low divergence (due to the scarcity of data in the
form of observed substitutions) and an overwhelming amount of
alignment error at high divergence (due to the large number of indel
events).

@@

A number of empirical analyses have established that errors in gene
sequencing, annotation and alignment can contribute to errors in
downstream evolutionary analyses such as phylogeny inference
\citep{Wong2008Alignment} and estimates of positive selection
\citep{Schneider2009Estimates,Markova-Raina2011High}. Most recently,
Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High} showed
that the detection of positively-selected sites and genes in \Dr
genomes is highly sensitive to aligner choice, with PRANK's codon
model \citep{Loytynoja2008PhylogenyAware} consistently producing
alignments with the lowest amount of positive selection. Still,
according to the authors' manual inspection of alignments, even
positively-selected sites identified with PRANK alignments contained a
sizable proportion of apparent false positives.

A limitation of the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
used their expected general effect of alignment error (an increase in
false positives due to misalignment of non-homologous codons) as a
proxy by which to compare different methods, allowing for the
conclusion that PRANK was the least error-prone aligner in their
analysis. However, the absolute number of false positives remained
uncertain and there was the possibility of conflating multiple sources
of error: in addition to alignment error, the authors noted that gene
mis-annotation was responsible for many apparent false positives, and
there is also an expected error rate from the likelihood inference
method itself. This limitation leaves important and interesting
questions, regarding the nature of alignment error and its
quantitative impact on the detection of positive selection unanswered
by empirical studies.

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. \textcolor{red}{GUIDANCE, which measures
  the robustness of alignment regions to perturbations in the guide
  tree used for progressive alignment, has also been proposed as an
  alignment confidence score
  \citep{Penn2010Alignment}}. Unfortunately, despite their widespread
use, the impact of the many available \textcolor{red}{alignment
  scoring and filtering} methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. With the application of published
filtering methods to alignments before testing for positive selection
becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach is similar to that of Anisimova et al.
\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
alignments to evaluate methods for detecting \sw positive selection,
considering each site to be an independent hypothesis test producing a
positive or negative result. Our experimental design also shares some
features with Fletcher and Yang's \citeyearpar{Fletcher2010Effect}
investigation of the performance of branch-site method for detecting
genewise positive selection under alignment error. Like Fletcher and
Yang \citeyearpar{Fletcher2010Effect}, we simulate codon alignments
with indels and calculate inferred alignments using a small but
diverse sample of aligners. Unlike their study, here we focus on \sw
detection of positive selection occurring throughout a phylogeny (as
opposed to genewise detection of selection acting at specific
branches), we explore a wider range of plausible tree sizes, indel
rates and divergence levels, and we evaluate the impact of a number of
alignment filtering methods on the \sw analysis. Given the recent
proliferation of completed and planned vertebrate genomes, we pay
special attention to choosing simulation parameters and analysis
methods similar to those commonly encountered in the \sw analysis of
vertebrate gene families.

@@

An overview of the simulation parameters used in this study can be
found in Table \ref{table_1}. Three rooted trees were used to guide
the simulation of protein-coding DNA alignments: the artificial
6-taxon tree used by Anisimova et al.
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate $\beta$-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution} and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure \ref{fig_1}, were scaled to
comparable divergence levels by normalizing their mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.05 and 2.0 synonymous substitutions per synonymous site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table
\ref{table_2}).
\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.5]{fig1.pdf}
\includegraphics[scale=0.8]{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis. The
  original scale for each tree is indicated by a scale bar, but trees
  were scaled to equal mean path length (MPL) divergence levels for
  simulation. (a) A 6-taxon artificial tree used in previous
  simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
  estimated from $\beta$-globin genes of 17 vertebrates and used in
  previous empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
  44-species tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes indicated by
  arrows were used as the reference species when comparing the true
  and inferred alignment (see Methods).}
\label{fig_1}
\end{figure}

The Indelible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons and $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4.
Indel lengths were drawn from a discretized
power-law distribution with an exponential decay parameter of 1.8 and
a maximum value of 40, yielding a mean indel length of 3.33 codons and
standard deviation of 5.51 codons. This power-law model of indel
lengths is well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and manual
inspection of alignments from a range of parameter values identified
the chosen model parameters as resulting in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was varied between 0 and 0.2 indel events per substitution
per site.

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(\citeyearpar{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were scaled before
simulation to correct for the difference between our definition of
branch lengths as the number of synonymous substitutions per
synonymous site ($dS$) and Indelible's interpretation of branch length
as the average number of substitutions per codon ($t$)
\citep{Fletcher2010Effect}. They are related approximately by
$t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
proportion of nonsynonymous and synonymous sites and $\bar{\omg}$ is
the mean $\omg$ across all sites. $S$ is approximately 0.3 when
$\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio
for our chosen distribution is 0.277, yielding a $dS$-to-$t$
conversion factor of 1.48 for all simulations performed.

@@

Alignments were filtered by masking out residues based on the output
of three alignment scoring methods: Gblocks conserved blocks
\citep{Castresana2000Selection}, T-Coffee consistency scores
\citep{Notredame2000TCoffee,Notredame2003Using}, and GUIDANCE
alignment confidence scores \citep{Penn2010Alignment}. Gblocks, which
identifies entire alignment columns as conserved or not conserved, was
run using an increased gap tolerance and a reduced minimum block
length in order to reduce the amount of each alignment removed
(command-line parameters {\em b5=a} and {\em b4=3}), and all residues
from any columns not within an identified conserved block were masked
with $N$s.

\textcolor{red}{In contrast to Gblocks, GUIDANCE and T-Coffee produce
  individual scores for each residue,} \textcolor{red}{ allowing for
  individual residues to be masked instead of entire columns. GUIDANCE
  generates many replicate alignments, each using a slightly perturbed
  guide tree, and assigns to each residue a score from 0 to 1 based on
  how consistently it was placed in the replicate alignments. We ran
  GUIDANCE with 100 replicates using MAFFT as the aligner, except when
  \prankc was the original aligner, in which case 30 replicate
  alignments were produced using \pranka. T-Coffee calculates the
  consistency between an input multiple alignment and independently
  calculated pairwise alignments \citep{Notredame2003Using}, rounding
  and normalizing residue scores into integers between 0 and
  9. T-Coffee was run using its default settings and the {\em
    evaluate\_mode --output=score\_ascii} command-line parameters to
  output alignment scores.}

\textcolor{red}{To filter alignments based on these residue-wise
  scores, a cutoff threshold was chosen for each method (0.5 for
  GUIDANCE and 5 for T-Coffee) and residues equal to or below that
  threshold were masked. On a per-alignment basis, if the default
  threshold caused greater than 50\% of residues to be masked, then
  the threshold was relaxed to the highest value for which at least
  50\% of residues remained unmasked. We found this adjustment
  necessary because the scores from GUIDANCE and T-Coffee were
  strongly affected by the simulation conditions, with much lower
  average scores at higher indel rates and divergences. Requiring at
  least 50\% of residues to remain unmasked ensured that enough data
  were available for meaningful evolutionary analysis, mimicking
  typical treatment of real data sets.}

Two unrealistic but informative datasets were produced to serve as controls. First,
the true simulated alignment was included in order to evaluate the \sw
performance without any alignment error. Second, an additional
filtering method was constructed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue reflecting how correctly
is has been placed in the inferred alignment. The approach taken was
to calculate, for each residue, the branch length of the correct
sub-tree (defined as the sub-tree connecting all sequences to which
the current residue was correctly aligned) divided by the branch
length of the total aligned sub-tree (defined as the sub-tree
connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
reflects the expectation that correctly-aligned evolutionary branch
length is the main source of information from which \sw inference
methods derive their power. We refer to this method as the 'optimal'
filtering method. Scores were handled in a manner similar to GUIDANCE and
T-Coffee, using a score threshold of 0.5.

@@

\Sw estimates of selective pressures were calculated using
maximum-likelihood methods implemented in the Phylogenetic Analysis by
Maximum Likelihood (PAML; \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
(SLR; \citealt{Massingham2005Detecting}) software packages.

@@

SLR implements a method specifically designed for \sw estimates which
has been shown in simulations to perform as well as or better than
PAML's \sw random sites models \citep{Massingham2005Detecting}. SLR
models codon evolution as a continuous-time Markov process where
substitutions at one site are independent of substitutions at all
other sites. No assumptions are made regarding the
distribution of $\omg$ ratios within the alignment. The value of
$\omg$ is considered to be an independent parameter at each site:
after first optimizing shared parameters using the whole alignment,
SLR uses the shared parameters and the data at each alignment site to
calculate a \sw statistic for non-neutral evolution. This statistic is
based on a likelihood-ratio test where the null model is neutral
evolution ($\omg=1$) and the alternative model is either purifying or
positive selection ($\omg<1$ or $\omg>1$, respectively). The raw
statistic measures the strength of evidence for non-neutral evolution
at each site; following Massingham and Goldman
\citeyearpar{Massingham2005Detecting} we use a signed version of the
SLR statistic (created by negating the statistic for sites with
$\omg<1$) as the test statistic for positive selection.

@@

In order to compare \sw estimates from different alignments, a single
sequence from each tree was chosen as the reference (arrows, Figure
\ref{fig_1}) and all \sw statistics were mapped from alignment columns
to sequence positions in the reference sequence. This approach bears
some resemblance to the process of mapping alignment-based
evolutionary estimates onto a single member of the alignment for
further analysis and integration with other genome-referenced data (as
is often done, for example, using mammalian alignments and a human
reference). As a result of this reference sequence based mapping,
sites which were deleted in the reference sequence or inserted in a
lineage not ancestral to the reference were \textcolor{red}{not
  included in the final performance analysis.}

To evaluate the power and error rates that might be achieved in
real-world data analysis, the recommended cutoff thresholds for PAML's
Bayesian posterior probabilities and the SLR statistic were used to
identify positively selected sites. A posterior probability threshold
of 0.95 was used for PAML \citep{Yang2005Bayes} and a threshold of 3.84,
the 95\% critical value of the $\chi^2$ distribution with 1 degree
of freedom, was used for SLR \citep{Massingham2005Detecting}. Sites were
compared to their true simulated state (e.g. positively-selected or
non-positively selected) in order to identify correct and incorrect
inferences, and from these classifications we calculated the false
positive rate (FPR, defined as the proportion of all sites with true
$\omg<1$ falsely identified as positively selected) and true positive rate
(TPR, defined as the proportion of all sites with true $\omg>1$ correctly
identified as positively selected).

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.75]{fig2.pdf}
\includegraphics[scale=1.25]{fig2.pdf}
\end{center}
\caption{Alignments were simulated without indels for three tree
  shapes and analyzed with SLR, \meight, or \mtwo. Fifty replicate
  alignments were simulated for each data point. The performance of
  each analysis method, as measured by four summary statistics, is
  plotted as a function of the mean path length (MPL) divergence. From
  top to bottom: true positive rate (TPR) at the recommended cutoff
  threshold (0.95 for PAML and 3.84 for SLR); false positive rate
  (FPR) at the recommended cutoff threshold; TPR at a 1\% FPR
  threshold; Pearson's correlation coefficient between the true and
  inferred sitewise $\omg$.}
\label{fig_2}
\end{figure}

\section*{Results and Discussion}
\subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
We first evaluated the ability of three \sw methods, \mtwo,
\meight and SLR, to accurately estimate \sw $\omg$ values and to
detect positive selection under a range of tree lengths in the absence
of alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw $\omg$
correlation over a range of MPLs for each of the three simulation
trees.
